---
title: "Comparing SilviMetric and FUSION outputs"
author: "Robert J. McGaughey"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(mapview)
```

## Setup

```{r}
HAGVRTfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/plumas_vrt_tifs/"
HAGNNfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/plumas_nn_tifs/"
HAGFUSIONfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/plumas_normalized_tifs/"
FUSIONfolder <- "H:/FUSIONTestMetrics/Products_FUSIONTestMetrics_2024-05-16/FINAL_FUSIONTestMetrics_2024-05-16/Metrics_30METERS/"

SMfile <- "m_Z_max.tif"
#SMfile <- "m_Z_mean.tif"
FUSIONfile <- "elev_max_2plus_30METERS.img"
```

## Read raster data and display extent and summary info

```{r echo=FALSE, results='hold'}
HAGVRTrast <- rast(paste0(HAGVRTfolder, SMfile))
HAGNNrast <- rast(paste0(HAGNNfolder, SMfile))
HAGFUSIONrast <- rast(paste0(HAGFUSIONfolder, SMfile))
FUSIONrast <- rast(paste0(FUSIONfolder, FUSIONfile))

cat("FUSION raster:    ncol=", ncol(FUSIONrast), "   nrow=",nrow(FUSIONrast), "   cells=", ncol(FUSIONrast) * nrow(FUSIONrast), "\n")
cat("HAGVRT raster:    ncol=", ncol(HAGVRTrast), "   nrow=",nrow(HAGVRTrast), "   cells=", ncol(HAGVRTrast) * nrow(HAGVRTrast), "\n")
cat("HAGNN raster:     ncol=", ncol(HAGNNrast), "   nrow=",nrow(HAGNNrast), "   cells=", ncol(HAGNNrast) * nrow(HAGNNrast), "\n")
cat("HAGFUSION raster: ncol=", ncol(HAGFUSIONrast), "   nrow=",nrow(HAGFUSIONrast), "   cells=", ncol(HAGFUSIONrast) * nrow(HAGFUSIONrast), "\n")

cat("FUSION raster:    ")
ext(FUSIONrast)
cat("HAGVRT raster:    ")
ext(HAGVRTrast)
cat("HAGNN raster:     ")
ext(HAGNNrast)
cat("HAGFUSION raster: ")
ext(HAGFUSIONrast)
```

## Adjust extent

The rasters produced when using PDAL's hag_nn are different from those produced using other hag filters.
I don't know for sure why but suspect it has to do with the extent of ground points compared to the extent
of non-ground points.

The HAGNN raster requires an extra operation to make it match the other rasters.

FUSION rasters are assigned a srs using an ESRI's projection file. The format of these files differs from 
the format used for the srs in the point cloud files so SilviMetric's srs doesn't exactly match FUSION's.
To overcome this, I forced the srs for FUSION rasters to match those from SilviMetric.

```{r echo=FALSE, results='hold'}
tHAGVRTrast <- trim(HAGVRTrast)
tHAGNNrast <- trim(HAGNNrast)   # different min Y///smaller by 1 cell
tHAGNNrast <- crop(tHAGNNrast, tHAGVRTrast)
tHAGFUSIONrast <- trim(HAGFUSIONrast)
tFUSIONrast <- trim(FUSIONrast)
cat("adjusted FUSION raster:    ")
ext(tFUSIONrast)
cat("adjusted HAGVRT raster:    ")
ext(tHAGVRTrast)
cat("adjusted HAGNN raster:     ")
ext(tHAGNNrast)
cat("adjusted HAGFUSION raster: ")
ext(tHAGFUSIONrast)

# force FUSION srs to match
crs(tFUSIONrast) <- crs(tHAGFUSIONrast)
```

## Extent and summary information for adjusted rasters

```{r echo=FALSE, results='hold'}
cat("FUSION output:    ncol=", ncol(tFUSIONrast), "   nrow=",nrow(tFUSIONrast), "   cells=", ncol(tFUSIONrast) * nrow(tFUSIONrast), "\n")
cat("HAGVRT output:    ncol=", ncol(tHAGVRTrast), "   nrow=",nrow(tHAGVRTrast), "   cells=", ncol(tHAGVRTrast) * nrow(tHAGVRTrast), "\n")
cat("HAGNN output:     ncol=", ncol(tHAGNNrast), "   nrow=",nrow(tHAGNNrast), "   cells=", ncol(tHAGNNrast) * nrow(tHAGNNrast), "\n")
cat("HAGFUSION output: ncol=", ncol(tHAGFUSIONrast), "   nrow=",nrow(tHAGFUSIONrast), "   cells=", ncol(tHAGFUSIONrast) * nrow(tHAGFUSIONrast), "\n")

cat("FUSION raster:\n")
summary(tFUSIONrast)
cat("HAGVRT raster:\n")
summary(tHAGVRTrast)
cat("HAGNN raster:\n")
summary(tHAGNNrast)
cat("HAGFUSION raster:\n")
summary(tHAGFUSIONrast)
```

## Compare raster values for maximum height

This metric (maximum HAG) doesn't involve any calculation so it basically tests that FUSION and SilviMetric
are using the same HAG values and same point for the cells.

This is a plot of the FUSION output for maximum height. 

```{r echo=FALSE}
plot(FUSIONrast, main = "FUSION -- maxHt")
```

The first test uses HAG computed using the VRT with PDAL's [hag_dem](https://pdal.io/en/stable/stages/filters.hag_dem.html)
filter. The interpolation problem with PDAL mentioned above leads 
to small differences between HAG compute by FUSION and using the VRT. In general, these differences are 
largest at the edge of the coverage area but there are patterns related to topogrphic features visible
in the difference raster.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGVRTrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- maxHt", xlab = "HAG computed using VRT")
plot(r, main = "FUSION minus SilviMetric -- maxHt")
par(mfrow = c(1, 1))
```

The second test uses HAG computed using PDAL's [hag_nn](https://pdal.io/en/stable/stages/filters.hag_nn.html)
filter. This produces similar differences compared to the VRT method. Again, differences are largest at the edge 
of the coverage area. Interesting, topographic patterns are not reflected in the differences but are when using the
VRT approach.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGNNrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- maxHt", xlab = "HAG computed using hag_nn")
plot(r, main = "FUSION minus SilviMetric -- maxHt")
par(mfrow = c(1, 1))
```

The final test used FUSION to compute HAG for the point tiles. Then PDAL to convert the tiles back to COPC 
format (FUSION doesn't write COPC format). The normalized point tiles were then used with SilviMetric (so no HAG computation was needed). This test, compared to the first test, highlights the difference in the HAG values
computed in FUSION and PDAL. The maximum heights for cells are nearly identical with differences attributable 
to numeric precision used for the HAG values (FUSION carries more significant digits...not necessarily more
accurate values).

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGFUSIONrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- maxHt", xlab = "HAG computed using points normalized by FUSION")
plot(r, main = "FUSION minus SilviMetric -- maxHt")
par(mfrow = c(1, 1))
```

I suspect that differences in the third test also result from the rules used to select point for a cell. FUSION
does not include points that excatly fall on the top and right edges of a cell whereas, PDAL includes these points.
While this does produce slightly different values for metrics, I don't think it affect the utility of the metrics.
It is debatable which method is more "correct" but I don't expect to see large difference in metrics over large
areas because of this difference...only for scattered individual cells.

## Compare raster values for mean height

This metric (mean HAG) involves calculation so it tests that FUSION and SilviMetric are using the same points
and calculation methods. 

```{r echo=FALSE, results='hold'}
SMfile <- "m_Z_mean.tif"
FUSIONfile <- "elev_ave_2plus_30METERS.img"

HAGVRTrast <- rast(paste0(HAGVRTfolder, SMfile))
HAGNNrast <- rast(paste0(HAGNNfolder, SMfile))
HAGFUSIONrast <- rast(paste0(HAGFUSIONfolder, SMfile))
FUSIONrast <- rast(paste0(FUSIONfolder, FUSIONfile))

tHAGVRTrast <- trim(HAGVRTrast)
tHAGNNrast <- trim(HAGNNrast)   # different min Y///smaller by 1 cell
tHAGNNrast <- crop(tHAGNNrast, tHAGVRTrast)
tHAGFUSIONrast <- trim(HAGFUSIONrast)
tFUSIONrast <- trim(FUSIONrast)

# force FUSION srs to match
crs(tFUSIONrast) <- crs(tHAGFUSIONrast)
```

This is a plot of the FUSION output for average height. 

```{r echo=FALSE}
plot(FUSIONrast, main = "FUSION -- aveHt")
```

This is the difference using HAG computed using a VRT with FUSION-derived DEM tiles.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGVRTrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- aveHt", xlab = "HAG computed using VRT")
plot(r, main = "FUSION minus SilviMetric -- aveHt")
par(mfrow = c(1, 1))
```

This is the difference using HAG computed using dem_nn.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGNNrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- aveHt", xlab = "HAG computed using hag_nn")
plot(r, main = "FUSION minus SilviMetric -- aveHt")
par(mfrow = c(1, 1))
```

This is the difference using HAG computed by FUSION.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGFUSIONrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- aveHt", xlab = "HAG computed using points normalized by FUSION")
plot(r, main = "FUSION minus SilviMetric -- aveHt")
par(mfrow = c(1, 1))
```

As with the comparison of maximum height values, the smallest differences occur when FUSION was used to normalize
the point data. For this metric, the HAG and the set of points due to inclusion of points on the right and top edges of cells affect the comparison.

## CHM comparison

SilviMetric was used to produce a CHM using 1.5m resolution. HAG was computed using the FUSION-derived DEM 
tiles and below ground points were dropped from the data. As with the metric rasters, some cropping of the 
FUSION output was needed to match the extent of the SilviMetric CHM. FUSION's normal logic fills holes
(areas with no points) using bilinear interpolation. This was turned off when creating this CHM.

**FUSION's alignment for surfaces is different from the alignment for raster layers. For the CHM, the cell
edges are a multiple of the cell size rather than the cell centers. This originated due to differences
between lattice and raster (or GRID) formats and alignment. A version of SilviMetric that matched this alignment was used to create the SilviMetric CHM.**

Looking closely at the following plots for my first run, I didn't see many empty (red) cells (NODATA or NA values for cells) in the FUSION output but did see some in the SilviMetric output. FUSION's behavior (CanopyModel) is 
to replace any valid negative point height value with 0.0. The logic in my build_pipeline() function drops 
points if there HAG is below 0.0. I set the min_HAG value to -100.0 and added an assign filter to set the Z 
(HAG) value for points below 0.0 to 0.0 and the empty cells were eliminated.

As with the HAG metrics, I suspect the interpolation process is causing the bulk of the difference. The remainder
is due to the right and top edge points in cells in SilviMetric but these should be small differences.

```{r echo=FALSE, results='hold'}
# compare CHMs
# NOTE: need to use PixelIsArea alignment in SilviMetric to align with FUSION CHM
# file name from FUSION hints at hole filling but this was disabled when creating the CHM
CHMVRTfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/plumas_CHM_vrt_tifs/"
SMfile <- "m_Z_max.tif"
r <- rast(paste0(CHMVRTfolder, SMfile))

plot(r, main = "SilviMetric -- CHM with VRT", colNA = "red")

FUSIONfolder <- "H:/FUSIONTestMetrics/Products_FUSIONTestMetrics_2025-02-04/FINAL_FUSIONTestMetrics_2025-02-04/CanopyHeight_1p5METERS/"
FUSIONfile <- "CHM_filled_not_smoothed_1p5METERS.img"
fr <- rast(paste0(FUSIONfolder, FUSIONfile))

plot(fr, main = "FUSION -- CHM", colNA = "red")

crs(fr) <- crs(r)

cat("SilviMetric raster:\n")
summary(r)

cat("\n\nFUSION raster:\n")
summary(fr)

tfr <- crop(fr, r)
diff <- tfr - r

#par(mfrow = c(2, 1))
cat("\n\nDifference raster:\n")
summary(diff)
hist(diff, nclass= 100, main = "FUSION minus SilviMetric -- CHM", xlab = "HAG computed using VRT")
plot(diff, main = "FUSION minus SilviMetric -- CHM")
par(mfrow = c(1, 1))

```

## SilviMetric CHM using fusion-normalized points

For completeness, I created another CHM using SilviMetric using the points normalized by FUSION. Differences are
slightly smaller than the CHM created using VRT for normalization. Both CHMs are adequate for all applications.

```{r echo=FALSE, results='hold'}
# compare CHMs
# NOTE: need to use pixel-is-cell alignment in SilviMetric to align with FUSION CHM
# file name from FUSION hints at hole filling but this was disabled when creating the CHM
CHMVRTfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/plumas_CHM_normalized_tifs/"
SMfile <- "m_Z_max.tif"
r <- rast(paste0(CHMVRTfolder, SMfile))

plot(r, main = "SilviMetric -- CHM with FUSION-normalized points", colNA = "red")

FUSIONfolder <- "H:/FUSIONTestMetrics/Products_FUSIONTestMetrics_2025-02-04/FINAL_FUSIONTestMetrics_2025-02-04/CanopyHeight_1p5METERS/"
FUSIONfile <- "CHM_filled_not_smoothed_1p5METERS.img"
fr <- rast(paste0(FUSIONfolder, FUSIONfile))

plot(fr, main = "FUSION -- CHM", colNA = "red")

crs(fr) <- crs(r)

cat("SilviMetric raster:\n")
summary(r)

cat("\n\nFUSION raster:\n")
summary(fr)

tfr <- crop(fr, r)
diff <- tfr - r

#par(mfrow = c(2, 1))
cat("\n\nDifference raster:\n")
summary(diff)
hist(diff, nclass= 100, main = "FUSION minus SilviMetric -- CHM", xlab = "HAG from FUSION-normalized points")
plot(diff, main = "FUSION minus SilviMetric -- CHM")
par(mfrow = c(1, 1))

```
