---
title: "Comparing SilviMetric and FUSION outputs for small test area"
author: "Robert J. McGaughey"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(terra)
library(sf)
library(mapview)
```

## Setup

```{r}
HAGVRTfolder <- "C:/Users/bmcgaughey/SilviMetricTesting/TestOutput/PlumasSmallArea_vrt_tifs/"
FUSIONfolder <- "H:/FUSIONTestMetrics/SmallArea/Products_SilviMetricTesting_2025-04-11/FINAL_SilviMetricTesting_2025-04-11/Metrics_30METERS/"

SMfile <- "m_Z_max.tif"
FUSIONfile <- "elev_max_2plus_30METERS.img"
```

## Read raster data and display extent and summary info

```{r echo=FALSE, results='hold'}
HAGVRTrast <- rast(paste0(HAGVRTfolder, SMfile))
FUSIONrast <- rast(paste0(FUSIONfolder, FUSIONfile))

cat("FUSION raster:    ncol=", ncol(FUSIONrast), "   nrow=",nrow(FUSIONrast), "   cells=", ncol(FUSIONrast) * nrow(FUSIONrast), "\n")
cat("HAGVRT raster:    ncol=", ncol(HAGVRTrast), "   nrow=",nrow(HAGVRTrast), "   cells=", ncol(HAGVRTrast) * nrow(HAGVRTrast), "\n")

cat("FUSION raster:    ")
ext(FUSIONrast)
cat("HAGVRT raster:    ")
ext(HAGVRTrast)
```

## Adjust extent

The rasters produced when using PDAL's hag_nn are different from those produced using other hag filters.
I don't know for sure why but suspect it has to do with the extent of ground points compared to the extent
of non-ground points.

The HAGNN raster requires an extra operation to make it match the other rasters.

FUSION rasters are assigned a srs using an ESRI's projection file. The format of these files differs from 
the format used for the srs in the point cloud files so SilviMetric's srs doesn't exactly match FUSION's.
To overcome this, I forced the srs for FUSION rasters to match those from SilviMetric.

```{r echo=FALSE, results='hold'}
tHAGVRTrast <- trim(HAGVRTrast)
tFUSIONrast <- trim(FUSIONrast)
cat("adjusted FUSION raster:    ")
ext(tFUSIONrast)
cat("adjusted HAGVRT raster:    ")
ext(tHAGVRTrast)

# force FUSION srs to match
crs(tFUSIONrast) <- crs(tHAGVRTrast)
```

## Extent and summary information for adjusted rasters

```{r echo=FALSE, results='hold'}
cat("FUSION output:    ncol=", ncol(tFUSIONrast), "   nrow=",nrow(tFUSIONrast), "   cells=", ncol(tFUSIONrast) * nrow(tFUSIONrast), "\n")
cat("HAGVRT output:    ncol=", ncol(tHAGVRTrast), "   nrow=",nrow(tHAGVRTrast), "   cells=", ncol(tHAGVRTrast) * nrow(tHAGVRTrast), "\n")

cat("FUSION raster:\n")
summary(tFUSIONrast)
cat("HAGVRT raster:\n")
summary(tHAGVRTrast)
```

## Compare raster values for maximum height

This metric (maximum HAG) doesn't involve any calculation so it basically tests that FUSION and SilviMetric
are using the same HAG values and same point for the cells.

This is a plot of the FUSION output for maximum height. 

```{r echo=FALSE}
plot(FUSIONrast, main = "FUSION -- maxHt")
```

The first test uses HAG computed using the VRT with PDAL's [hag_dem](https://pdal.io/en/stable/stages/filters.hag_dem.html)
filter. The interpolation problem with PDAL mentioned above leads 
to small differences between HAG compute by FUSION and using the VRT. In general, these differences are 
largest at the edge of the coverage area but there are patterns related to topogrphic features visible
in the difference raster.

```{r echo=FALSE, results='hold'}
#par(mfrow = c(2, 1))
r <- tFUSIONrast - tHAGVRTrast
summary(r)
hist(r, nclass= 100, main = "FUSION minus SilviMetric -- maxHt", xlab = "HAG computed using VRT")
plot(r, main = "FUSION minus SilviMetric -- maxHt")
par(mfrow = c(1, 1))
```
